---
title: "Human PDGFRb+ : pathway characterization of mesenchymal NKD2+ cells"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

## Setting the environment

### Internal variables
```{r}
set.seed(1234)
OUTDIR <- "./human_PDGFRbpositive_02_pathway_output/";
if(!dir.exists(OUTDIR)) dir.create(OUTDIR, recursive = TRUE);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png', 'pdf'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);

# If already exists, clean dirs?
clean_dirs <- TRUE
if(clean_dirs) {
	unlink(list.files(OUTDIR, full.names=TRUE, recursive = TRUE))
}
```

### Load libraries
Libraries needed for the analysis
```{r}
suppressPackageStartupMessages(library("Matrix"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("SingleCellExperiment"))
suppressPackageStartupMessages(library("scran"))
suppressPackageStartupMessages(library("cowplot"))
suppressPackageStartupMessages(library("ComplexHeatmap"))
suppressPackageStartupMessages(library("dorothea"))
suppressPackageStartupMessages(library("progeny"))
suppressPackageStartupMessages(library("ggrepel"))
suppressPackageStartupMessages(library("gridExtra"))
```

### Functions

An enhanced version of Seurat's `DoHeatmap` to highlight scaled features in single-cell dataset.
```{r}
# Enhanced DoHeatmap function
DoHeatmap <- function(mat,cl,origin, 
		       row_names_size=14, 
		       dat="scale", column_title_size=14) {
  
  gg_color_hue <- function(n) {
    hues = seq(15, 375, length = n + 1)
    hcl(h = hues, l = 65, c = 100)[1:n]
  }
  
  gg_color_hue2 <- function(n) {
    hues = seq(15, 375, length = n + 1)
    hcl(h = hues, l = 35, c = 100)[1:n]
  }
  
  gg_color_hue3 <- function(n) {
    hues = seq(15, 375, length = n + 1)
    hcl(h = hues, l = 95, c = 100)[1:n]
  }
  
  f1 <-  circlize::colorRamp2(c(-2,0,+2), c("purple", "black", "yellow"))
 
  # Reorder
  ord <- order(as.numeric(cl), decreasing = FALSE)
  mat <- mat[,ord]
  
  cl <- cl[ord]
  cl.cols <- setNames(gg_color_hue(length(unique(cl))),levels((cl)))
  
  origin <- origin[ord]
  if(length(unique(origin)) > 2) {
	  origin.cols <- setNames(gg_color_hue3(length(unique(origin))),
                          unique(as.character(origin)))
  } else if (length(unique(origin))==2) {
	 origin.cols <- setNames(c("red","grey"), unique(as.character(origin)))
  }
		
   hc <- HeatmapAnnotation(df=data.frame("cluster"=as.character(cl), "origin"=origin, row.names=colnames(mat)),
                          col = list("cluster"=cl.cols, "origin"=origin.cols),
                          show_annotation_name = TRUE,annotation_name_side = "left",
                          show_legend = c(FALSE,TRUE),
			  annotation_legend_param = list("cluster"=list(nrow=1),
							 "origin"=list(nrow=1, 
								       title_gp = gpar(fontsize=16),
								       labels_gp = gpar(fontsize=14))
							 )
			  )
  # Choose main legend scale
  if(dat=="scale") {
  	f1 <-  circlize::colorRamp2(c(-2,0,+2), c("purple", "black", "yellow"))
  } else if(dat=="data") {
  	f1 <-  circlize::colorRamp2(c(min(as.vector(mat2)),
					max(as.vector(mat2))),
				      c("purple","yellow"))
  } else {
	  stop("ERROR: data type is not supported")
  }
  

   hp <- Heatmap(mat, cluster_rows = FALSE, cluster_columns = FALSE,col = f1,use_raster=FALSE,
                name="Activity",
		width = unit(28, "cm"),
		height = unit(8, "cm"),
                 top_annotation = hc,# bottom_annotation = hc,
                  column_split = factor(cl, levels=unique(cl)), column_title_rot=00, column_title_gp = gpar(fontsize=column_title_size), column_gap = unit(0.7, "mm"),
                heatmap_legend_param= list(legend_height = unit(4, "cm"),
					   legend_direction = "horizontal",
                                           title_gp=gpar(fontsize=16),
                                           labels_gp = gpar(fontsize = 15)),
                show_column_names = FALSE, row_names_side = "left",
                row_names_gp = gpar(fontsize=row_names_size))
  return(hp)
}

```

We also define a split-violin like plot to highlight differences between groups.
```{r splitVln}
# source: split-violin at https://stackoverflow.com/questions/35717353/split-violin-plot-with-ggplot2
# source: quantiles in split-vilin at https://stackoverflow.com/questions/47651868/split-violin-plot-with-ggplot2-with-quantiles/47652563#47652563
GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin,
  draw_group = function(self, data, ..., draw_quantiles = NULL) {
    # Original function by Jan Gleixner (@jan-glx)
    # Adjustments by Wouter van der Bijl (@Axeman)
    data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
    grp <- data[1, "group"]
    newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])
    if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
      stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <= 1))
      quantiles <- create_quantile_segment_frame(data, draw_quantiles, split = TRUE, grp = grp)
      aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
      aesthetics$alpha <- rep(1, nrow(quantiles))
      both <- cbind(quantiles, aesthetics)
      quantile_grob <- GeomPath$draw_panel(both, ...)
      ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
    }
    else {
      ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
    }
  }
)

create_quantile_segment_frame <- function(data, draw_quantiles, split = FALSE, grp = NULL) {
  dens <- cumsum(data$density) / sum(data$density)
  ecdf <- stats::approxfun(dens, data$y)
  ys <- ecdf(draw_quantiles)
  violin.xminvs <- (stats::approxfun(data$y, data$xminv))(ys)
  violin.xmaxvs <- (stats::approxfun(data$y, data$xmaxv))(ys)
  violin.xs <- (stats::approxfun(data$y, data$x))(ys)
  if (grp %% 2 == 0) {
    data.frame(
      x = ggplot2:::interleave(violin.xs, violin.xmaxvs),
      y = rep(ys, each = 2), group = rep(ys, each = 2)
    )
  } else {
    data.frame(
      x = ggplot2:::interleave(violin.xminvs, violin.xs),
      y = rep(ys, each = 2), group = rep(ys, each = 2)
    )
  }
}

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, position = position, 
        show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
```

An enhanced version of progenyScatter for visualization of responsive genes to pathway perturbation.
```{r}
progenyScatter2 <- function (df, weight_matrix, dfID = 1, weightID = 1, statName = "gene stats",
    verbose = FALSE)
{
    weight <- color <- ID <- NULL
    plot_list_contrasts <- list(0)
    for (i in 2:length(df[1, ])) {
        plot_list_pathways <- list(0)
        for (j in 2:length(weight_matrix[1, ])) {
            sub_df <- df[, c(dfID, i)]
            pathway_weights <- weight_matrix[, c(weightID, j)]
            names(sub_df) <- c("ID", "stat")
            minstat <- min(sub_df$stat)
            maxstat <- max(sub_df$stat)
            histo <- ggplot(sub_df, aes(x = stat, fill = "blue")) +
                geom_density() + coord_flip() + 
                geom_vline(xintercept = 0, linetype = "dotted") +
		scale_fill_manual(values = c("#00c5ff")) +
                xlim(minstat, maxstat) + theme_cowplot() + 
		theme(legend.position = "none",
                	axis.text.x = element_text(size=18, colour="white"), 
			axis.ticks.x = element_blank(),
			axis.title.y = element_blank(),
			# Otherwise both plots does not fit in same margin
			axis.title.x = element_text(size=18),
			axis.text.y = element_blank(),
			axis.ticks.y = element_blank(), 
			panel.grid.major = element_blank(),
			axis.line = element_line(colour="white"),
                panel.grid.minor = element_blank(),
		plot.background = element_rect(colour = "white", fill="white")
# 		plot.margin = unit(c(1,1,1.73,1), "cm")
		)
            names(pathway_weights) <- c("ID", "weight")
            pathway_weights <- pathway_weights[pathway_weights$weight !=
                0, ]
            percentile <- ecdf(sub_df$stat)
            sub_df <- merge(sub_df, pathway_weights, by = "ID")
            sub_df$color <- "3"
            sub_df[(sub_df$weight > 0 & sub_df$stat > 0), "color"] <- "1"
            sub_df[(sub_df$weight > 0 & sub_df$stat < 0), "color"] <- "2"
            sub_df[(sub_df$weight < 0 & sub_df$stat > 0), "color"] <- "2"
            sub_df[(sub_df$weight < 0 & sub_df$stat < 0), "color"] <- "1"
            sub_df[(percentile(sub_df$stat) < 0.95 & percentile(sub_df$stat) >
                0.05), 1] <- NA

	    maxwei <- max(sub_df$weight, na.rm=TRUE)
	    minwei <- min(sub_df$weight, na.rm=TRUE)
            if (verbose) {
                message(paste("weights of ", names(weight_matrix)[j],
                  sep = ""))
            }
            title <- paste("weights of ", names(weight_matrix)[j],
                sep = "")
            scatterplot <- ggplot(sub_df, aes(x = weight, y = stat,
                color = color)) + geom_point() + scale_colour_manual(values = c("red",
                "royalblue3", "grey")) + 
                geom_vline(xintercept = 0, linetype = "dotted") +
                geom_hline(yintercept = 0, linetype = "dotted") +
# 		geom_label_repel(aes(label = ID)) +
		# Best up-pos
		geom_label_repel(data=subset(sub_df, 
					     	stat >0 & weight > 0
					     ),
					     mapping=aes(label = ID),
				 xlim=c(0,maxwei),
				 ylim=c(0,maxstat),
				 seed=1234,
				 force=2,
				 segment.alpha=0.3,
				) +
		# Best dn-neg
		geom_label_repel(data=subset(sub_df, 
					     	stat < 0 & weight < 0
					     ),
					     mapping=aes(label = ID),
				 xlim=c(minwei, 0),
				 ylim=c(minstat, 0),
				 seed=1234,
				 force=3,
				 segment.alpha=0.3,
				) +
		# Best dn-pos
		geom_label_repel(data=subset(sub_df, 
					     	stat < 0 & weight > 0
					     ),
					     mapping=aes(label = ID),
				 xlim=c(0, maxwei),
				 ylim=c(minstat,0),
				 seed=1234,
				) +
		# Best up-neg
		geom_label_repel(data=subset(sub_df, 
					     	stat > 0 & weight < 0
					     ),
					     mapping=aes(label = ID),
				 xlim=c(minwei, 0),
				 ylim=c(0, maxstat),
				 seed=1234,
				) +
                ylim(minstat, maxstat) + theme_cowplot() + 
		theme(legend.position = "none",
		      axis.text = element_text(size=18),
		      axis.title = element_text(size=18),
			plot.background = element_rect(colour = "white", 
						       fill="white"),
		      ) +
                labs(x = title, y = statName)
            lay <- t(as.matrix(c(1, 1, 1, 1, 2)))
            gg <- arrangeGrob(scatterplot, histo, nrow = 1, ncol = 2,
                layout_matrix = lay)
            plot_list_pathways[[j - 1]] <- gg
        }
        names(plot_list_pathways) <- names(weight_matrix[, -weightID])
        plot_list_contrasts[[i - 1]] <- plot_list_pathways
    }
    return(plot_list_contrasts)
}
```

```{r}
# Short annotation
#' A handy function
#' x is a vector or elements
ann2acr <- function(x) {
	xx <- gsub("Fibroblasts?","Fib", x) 
	xx <- gsub("Myofibroblasts?","MF", xx)
	xx <- gsub("Pericytes","Pe", xx)
	xx <- gsub("Smooth Muscle Cells","SMC", xx)
	xx <- gsub("Mesangial Cells","Mesa", xx)
	xx <- gsub("^Vascular","v",xx)
	xx <- gsub(" ","", xx)
	return(xx)
}
```

ggplot2 theme with minimal elements and large fonts. Two colors for main conditions.
```{r}
theme_min <- theme_cowplot() +
	theme(
	plot.title = element_text(size=28),
	  plot.margin = unit(c(1,0.3,0.3,0.3), "cm"),
	  axis.text.x = element_text(size=24),
	  axis.text.y = element_text(size=24),
	  axis.title = element_text(size=22),
	  legend.text = element_text(size=22),
	  legend.title = element_blank(),
	  legend.position="top",
	)

kidneyf_col <- c("grey60", "red")
kidneyf_bg <- c("grey", "red")
```

## Load data
```{r}
sce <- readRDS("./human_PDGFRbpositive_01_create_sce_output/sce.rds")
```

## Prepare data for visualization

```{r}
MES <- sce[,colData(sce)$Annotation.Level.1=="Mesenchymal"]
dim(MES)
rownames(MES) <- rowData(MES)$Gene.Symbol
MES <- MES[!duplicated(rownames(MES)), ]
MES <- MES[, order(MES$Kidney.Function, decreasing=FALSE)]
```

```{r hp_PROGENy, fig.width=14, fig.height=6, dpi=300}
# By hand: we rename the last group, unknown mesenchymal
# to avoid overlap
cl <- ann2acr(MES$Annotation.Level.3)

hp <- DoHeatmap(mat=assay(altExp(MES, "pathway")), 
		cl=as.factor(cl),
	   origin=MES$Kidney.Function)
draw(hp, heatmap_legend_side="bottom", annotation_legend_side="bottom")
```

## Stratification of Mesenchymal cells by NKD2 expression
Stratify cells by NKD2 expression, but also retrieve only PDGFRa/b+ cells.
```{r}
colData(MES)$Strat <- NA
colData(MES)$Strat[assay(MES, "binary_lvl3")["NKD2",]==1] <- "NKD2_pos"
colData(MES)$Strat[assay(MES, "binary_lvl3")["NKD2",]==0] <- "NKD2_neg"

# We are only interested in Pericytes, Fibroblasts, Myofibroblasts
MES <- MES[, MES$Annotation.Level.2 %in% c("Fibroblasts","Myofibroblasts","Pericytes")]
# Report
table(colData(MES)$Strat)
```

## single-cell PROGENy estimations
We run a wilcox test to rank pathways based on differences between NKD2+ and NKD2-.
```{r NKD2_progeny_wilcox}
# Wilcox test to get the p-values
res <- findMarkers(assay(altExps(MES)$pathway), 
		   groups=colData(MES)$Strat, 
		   test.type="wilcox")

as.data.frame(res$NKD2_pos)
tags <- data.frame("pathway"=rownames(res$NKD2_pos),
		   "pathway_sign"=paste0("Top #",res$NKD2_pos$Top,
					 " : ", rownames(res$NKD2_pos))
# 					 " (FDR =", formatC(res$NKD2_pos$FDR, format = "e", digits = 2),")")
		   )
tags$pathway_sign <- factor(tags$pathway_sign, levels=tags$pathway_sign)

# Save results
write.table(as.data.frame(res$NKD2_pos), 
	    file=paste0(DATADIR,"/pooledNKD2_progeny_wilcox.csv"), sep=",", 
	    col.names=NA, row.names=TRUE, quote=FALSE)
```

Actually, we re-run the test for independent cell types to get an estimation of the difference in each one of them,
based on the differences between NKD2+ and NKD2-.
```{r NKD2strat_progeny_wilcox}
RES <- sapply(unique(colData(MES)$Annotation.Level.2), function(cell) {
		cell.idx <- which(colData(MES)$Annotation.Level.2==cell)
		findMarkers(assay(altExps(MES)$pathway)[, cell.idx], 
		   		groups=colData(MES)$Strat[cell.idx], 
		   		test.type="wilcox")
	    })
RES <- lapply(RES, function(z) z$NKD2_pos)
for (cl in names(RES)) {
		RES[[cl]]$pathway <- rownames(RES[[cl]])
		RES[[cl]]$CellType <- cl
} 
RES <- do.call("rbind", RES)
# Multiple testing correction (FDR)
RES$FDR <- p.adjust(RES$p.value, method="fdr")
RES <- as.data.frame(RES)
# Reorder rows,cols
rownames(RES) <- NULL
col1st <- c("CellType","pathway")
RES <- RES[, c(col1st, setdiff(colnames(RES), col1st))]
rm(col1st)
# Save results
write.table(RES, 
	    file=paste0(DATADIR,"/wilcox_significance_for_extended_data_Fig10c.csv"), sep=",", 
	    col.names=TRUE, row.names=FALSE, quote=FALSE)

# add tag for visualization
RES$tag <- sapply(RES$FDR, function(pval) {
    if(pval< 0.001) {
        txt <- "***"
    } else if (pval < 0.01) {
        txt <- "**"
    } else if (pval < 0.05) {
        txt <- "*"
    } else {
        txt <- "ns"
    }
    return(txt)
	    })
```

Prepare data for visualization (single-cell estimates) with significance as annotation.
```{r NKD2_progeny_dat}
dat <- reshape2::melt(assay(altExps(MES)$pathway))
colnames(dat) <- c("pathway", "cellID", "score")
dat$Strat <- ifelse(colData(MES)[dat$cellID, "Strat"]=="NKD2_pos", "NKD2+", "NKD2-")
dat$ann2 <- colData(MES)[dat$cellID, "Annotation.Level.2"]
dat$sann2 <- ann2acr(dat$ann2)
dat <- merge(dat, tags, by.x="pathway", by.y="pathway", all.x=TRUE, all.y=FALSE)

# Order levels (pathways) by significance
dat$pathway <- factor(as.character(dat$pathway),
		      levels=unique(as.character(dat$pathway[order(dat$pathway_sign)])))

# For the records, how many (n) cells in each group
ncells <- table(dat$sann2,dat$Strat)/length(unique(dat$pathway_sign))
print(ncells)
```

For annotating significance, we add some data
```{r}
dat$key <- paste0(dat$ann2,"_",dat$pathway) 
RES$key <- paste0(RES$CellType,"_",RES$pathway) 
dat <- merge(dat, RES[,c("FDR", "tag", "key")], 
	     by.x="key", by.y="key", all.x=TRUE, all.y=FALSE)
dat$maxscore <- max(dat$score)
# Only one tag per key2, that is, per group by pathway. 
# So plot is not overwritten per each cellID (observation)
dat$key2 <- paste0(dat$pathway_sign,"_", dat$sann2,"_",dat$maxscore)
dat$tag2 <- NA 
dat$tag2[!duplicated(dat$key2)] <- dat$tag[!duplicated(dat$key2)]
```

We visualize the differences between populations as a violin plot split by NKD2 expression.
```{r extended_data_Fig10c, fig.width=16, fig.height=14, dpi=300, warning=FALSE}
ncells_ann <- paste0("NKD2(-/+) cells","  ",
		     rownames(ncells)[1],"(", paste(ncells[1,c("NKD2-","NKD2+")], collapse="/"),")","  ",
		     rownames(ncells)[2],"(", paste(ncells[2,c("NKD2-","NKD2+")], collapse="/"),")","  ",
		     rownames(ncells)[3],"(", paste(ncells[3,c("NKD2-","NKD2+")], collapse="/"),")","  "
		)

ggplot(dat,
       aes(x=sann2, y=score, fill=Strat)) +
    geom_split_violin(draw_quantiles = 0.5) + facet_wrap(.~ pathway,nrow=2) + 
    geom_text(aes(x=sann2, y=maxscore, label=tag2),
	      position = position_dodge(width = 0),
	      hjust = 0.5, size = 7) +
scale_fill_manual(values=kidneyf_bg) +
scale_colour_manual(values=kidneyf_col) +
    xlab("") + ggtitle("PROGENy pathways") + ylab("Scaled score") +
    theme_min + 
    labs(caption = ncells_ann) + 
    theme(strip.text.x = element_text(size=28, colour="black",hjust=0),
	  plot.caption = element_text(size=30, colour="black", hjust=0)
	  ) 

```

We zoom-in in TGFb: the most differentially active pathway in NKD2+ cells.
```{r NKd2_TGFb_progeny, fig.width=6, fig.height=6, dpi=300}
ggplot(subset(dat, pathway=="TGFb"),
       aes(x=sann2, y=score, fill=Strat)) +
    geom_split_violin(draw_quantiles = 0.5) +
scale_fill_manual(values=kidneyf_bg) +
scale_colour_manual(values=kidneyf_col) +
    xlab("") + ggtitle("TGFb pathway") + ylab("Scaled score") +
    theme_min
```

> We could observe the higher activity of TGFb in NKD2+ cells 
among Fibroblasts, Myofibroblasts and Pericytes.

## PROGENy on contrasts (based on gene-wise permutation of rankings)

Signature extraction for NKD2+ MES
```{r}
# Wilcox test to get the p-values
res <- findMarkers(logcounts(MES), groups=colData(MES)$Strat, test.type="wilcox")

# Ad-hoc log-fold-change calculation to get the sign of dysregulation
lfc <- rowMeans(logcounts(MES)[,which(colData(MES)$Strat=="NKD2_pos")]) - 
	rowMeans(logcounts(MES)[,which(colData(MES)$Strat=="NKD2_neg")])

# Volcano plot
df <- data.frame(lfc=lfc[rownames(res$NKD2_pos)],
		 pval=res$NKD2_pos$p.value,
		 FDR=res$NKD2_pos$FDR,
		 row.names=rownames(res$NKD2_pos))

df <- df[which(rownames(df)!="NKD2"),]
df <- df[which(rownames(df)!=""),]
```

```{r}
signat <- setNames(-log10(df$pval) * sign(df$lfc), rownames(df))
signat <- na.omit(signat)
```

We first calculate zscores from the 
wilcox test for a gene-permutation PROGENy estimation.
```{r}
# src: gCMAP::zScores
zScores <- function(pval, direction=NULL, tails=2,limit=.Machine$double.xmin) {
  if( !is.null( limit ) ){
    pval[which(pval < limit )] <- limit ## set lower limit to avoid Inf/-Inf zscores
  }
  if( tails == 2){
    z <- qnorm( pval/2, lower.tail=FALSE )
  } else if( tails == 1){
    z <- qnorm(pval, lower.tail = FALSE)
  } else {
    stop( "Parameter 'tails' must be set to either 1 or 2.")
  }
  if ( !is.null( direction) ) {
    z <-  z * sign( direction )
  }
  z
}

zz <- zScores(df$pval, df$lfc, tails=2)
signat.df <- data.frame(ID=rownames(df), NKD2=zz)
```

We calculate the PROGENy estimations in the contrast.
```{r strat_progrun, fig.width=1.2, fig.height=4, dpi=300}
mod <- getModel(organism = "Human", top = 500)
mod_cols <- colnames(mod)
mod$ID <- rownames(mod)
mod <- mod[,c("ID",mod_cols)]


pathway_scores <- progenyPerm(df=signat.df,
		    weight_matrix=mod,
		    k=10000,
		    z_scores=TRUE)
rownames(pathway_scores)<- NULL

path_mat <- t(pathway_scores)
path_mat <- path_mat[order(path_mat[,1],decreasing=TRUE),,drop=FALSE]

pscore_max <- ceiling(max(abs(as.vector(pathway_scores))))
col_fun <- circlize::colorRamp2(c(pscore_max,
				  0,
				  -1*pscore_max),
				c("yellow","black","purple"))
hp <- Heatmap(path_mat, 
	      col=col_fun,
	name="Activity",
	cluster_rows=FALSE, cluster_columns=FALSE, row_names_gp = gpar(fontsize=0),
                heatmap_legend_param= list(legend_direction = "horizontal", 
					   legend_width = unit(2.5,"cm"),
					   at=seq(from=-1*pscore_max,
						  to=pscore_max,
						  length.out=5),
					   title_gp=gpar(fontsize=10),
					   labels_gp=gpar(fontsize=9)),
	cell_fun=function(j,i,x,y,width, height,fill) {
		grid.text(rownames(path_mat)[i], x, y, gp=gpar(fontsize=10, 
							       fontface="bold",
							       col=ifelse(abs(path_mat[i,j])>6.5, "black","white")))
	})
draw(hp, heatmap_legend_side="bottom")
```

```{r scatter_progeny_TGFb, fig.width=18, fig.height=7, dpi=300, warning=FALSE}
scat_plots <- progenyScatter2(df=signat.df, weight_matrix=mod, 
			     statName="z-score",verbose=FALSE)
 
plot(scat_plots[[1]]$`TGFb`)
```

> Run PROGENy on contrasts is in agreement with single-cell PROGENy estimates. 
TGFb is the most active pathway in NKD2+ mesenchymal cells.

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
