---
title: "Differential expression analysis in Myofibroblast with NKD2 perturbations"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

> Rafael kramann's lab has established one cell line from myofibroblasts from kidney bipsies with CKD.
Then this cell line was transformed with NKD-KO, which generated three different clones carrying NKD2-KO.
In addition, myofibroblasts were also transformed with NKD2- over-expression. These gene perturbations are
compared with their corresponding controls. 

Previous script ([00_EDA.md](./00_EDA.md)) shows that:

* Biological replicates cluster together.

* One KO clone presents much more differences to the other KO clones. 
  In addition, that clone is actually a clone with a fibrotic severe phenotype with empirical observation.
  While the other two clones cluster together with a less severe phenotype.  Thus we define 2 groups accordingly
  for the differential expression: clone 1+3 as KO-shallow and the single clone 2 as KO-severe.

## Set-up environment
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Define random seed for reproduciblity, and output folder.
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./01_DGE_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png', 'pdf'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);

# If already exists, clean dirs?
clean_dirs <- FALSE
if(clean_dirs) {
	unlink(list.files(OUTDIR, full.names=TRUE, recursive = TRUE))
}
```


### Load libraries
Essential R libraries for the analysis.
```{r}
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(GSEABase))
suppressPackageStartupMessages(library(fgsea))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(grid))
fontTXT <- "sans"
```

### Functions
```{r pval2txt}
pval2txt <- function(padj) {
padj_txt <- apply(padj, c(1,2), function(pval) {
	if(!is.na(pval)) {
		   if(pval< 0.001) {
			   txt <- "***"
		   } else if (pval < 0.01) {
			   txt <- "**"
		   } else if (pval < 0.05) {
			   txt <- "*"
		   } else {
			   txt <- "ns"
		   } 
	} else {
		txt <- "n/a"
	}
	return(txt)
})

return(padj_txt)
}
```

Some handy functions for visualization
Make a heatmap with text showing significant hits.
```{r make_hp}
### Make a heatmap of the GSEA results for the 3 contrasts
# NES is a matrix of NES
# PADJ is a matrix of strings for each cell
makeHP <- function(NES, TXT) {

hp <- Heatmap(NES, column_names_side="top", row_names_side="left",
	      name="Normalized Enrichment Score (NES)",
	      heatmap_legend_param=list(direction="horizontal",
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
	show_column_dend=FALSE, show_row_dend=FALSE,	 
 	column_names_rot=0,
	row_names_gp=gpar(fontsize=12),
    	row_names_max_width = max_text_width(
        			rownames(NES), 
        			gp = gpar(fontsize = 12)),
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		grid.text(TXT[i, j], x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(NES[i,j] < -2,
						 "white","black"))
			  )
	}
)
return(hp)
}
```

Handy function to make a volcano to highlight a set of genes in the landscape of differential expression.
```{r make_volcano}
make_volcano <- function(topTab, set) {

if(any(!set %in% topTab$ID)) {
	cat(paste("The following genes are not present in the gene expr matrix:","\n",
		  paste(setdiff(set, rownames(topTab)), collapse=","),
		  "\n"),
		  file=stdout())
	set <- intersect(set, rownames(topTab))
}

topTab$significant <- ifelse(topTab$adj.P.Val < 0.05,"FDR<0.05","n.s.")
topTab$show <- topTab$ID %in% set
 
lfc <- topTab$logFC
mxpval <- max(-log10(topTab$P.Value))

p = ggplot(topTab, aes(logFC, -log10(P.Value))) +
    geom_point(aes(col=significant)) +
    scale_color_manual(values=c("red", "grey"))
  
  
p <- p+geom_label_repel(data=topTab[topTab$show,],
                          aes(label=ID), family=fontTXT, size=4,
                          force=2,
			  max.iter=1e4,
# 			  xlim = c(ceiling(max(lfc))+1, ceiling(max(lfc))+2),
 			  xlim = c(floor(min(lfc))*1.5, floor(min(lfc)*1.2)),
# 			  xlim = c(min(lfc), 0),
                           nudge_y=0.05, 
			  direction = "y", 
			  segment.size = 0.3) + 
    coord_cartesian(xlim = c(floor(min(lfc)), 
			     ceiling(max(lfc))), 
		    clip = "off") +
    scale_y_continuous(position="right") + 
    theme_cowplot() +
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  legend.justification=c(1,0),
	  axis.text = element_text(family=fontTXT, size=20),
     	plot.margin = unit(c(0.3, 2.6, 1.0, 4.0), "cm") 
	  )
return(p)

}
```

## Load data and normalize

```{r}
### 1 Load data
# Source: https://bioconductor.org/packages/release/bioc/vignettes/tximport/inst/doc/tximport.html#limma-voom
load("../data/bulk-rnaseq/forJavier.RData")
# The sample IDs are stored in the folder from the file that was imported to R. We recall them
sname <- sapply(dirname(filenames),function(z) paste(strsplit(z, split="_")[[1]][1:2],collapse="_"))
# We get the read counts per sample
cnt <- txi$counts
# Rename samples
colnames(cnt) <- sname
# Get the biological conditions
group_str <- as.character(group)
group_str <- gsub("-","_", group_str)
group_str <- gsub("OverExpr","OE",group_str)
```

### Knock-out experiment
Define data experiment
```{r}
KO_idx <- grep("^KO", group_str)
KO_cnt <- cnt[, KO_idx]
KO_gr <- factor(group_str[KO_idx])
KO_gr <- relevel(KO_gr, ref="KO_ctrl")
```

```{r}
## 2 Create DGElist (edgeR package, followed by TMM default normalization method)
KO_y <- DGEList(counts = KO_cnt, group = KO_gr, genes = rownames(KO_cnt))
# Filtering
KO_keep <- filterByExpr(KO_y)
table(KO_keep)

# Apply filtering
KO_y <- KO_y[KO_keep, ]
KO_y <- calcNormFactors(KO_y)

## 3 Create experimental design
stopifnot(all(KO_gr == KO_y$samples$group))
KO_design <- model.matrix(~ 0 + KO_gr) # wo/ intercept
colnames(KO_design) <- gsub("^KO_gr", "", colnames(KO_design))
KO_v <- voom(KO_y, KO_design)

# Retrieve only gene symbols for downstream analysis
rownames(KO_v) <- sapply(rownames(KO_v), function(z) strsplit(z, split=";")[[1]][2])
```

### Over-expression experiment
Define data experiment
```{r}
OE_idx <- grep("^OE", group_str)
OE_cnt <- cnt[, OE_idx]
OE_gr <- factor(group_str[OE_idx])
OE_gr <- relevel(OE_gr, ref="OE_ctrl")
```

```{r}
## 2 Create DGElist (edgeR package, followed by TMM default normalization method)
OE_y <- DGEList(counts = OE_cnt, group = OE_gr, genes = rownames(OE_cnt))
# Filtering
OE_keep <- filterByExpr(OE_y)
table(OE_keep)

# Apply filtering
OE_y <- OE_y[OE_keep, ]
OE_y <- calcNormFactors(OE_y)

## 3 Create experimental design
stopifnot(all(OE_gr == OE_y$samples$group))
OE_design <- model.matrix(~ 0 + OE_gr) # wo/ intercept
colnames(OE_design) <- gsub("^OE_gr", "", colnames(OE_design))
OE_v <- voom(OE_y, OE_design)

# Retrieve only gene symbols for downstream analysis
rownames(OE_v) <- sapply(rownames(OE_v), function(z) strsplit(z, split=";")[[1]][2])
```

## Test for differential expression

### Knock-out experiment
```{r KO_contrast}
KO_fit <- lmFit(KO_v, KO_design)
KO_cont.mat <- makeContrasts("NKD2_KO_severe"=KO_1 - KO_ctrl,
			  "NKD2_KO_shallow"=(KO_2 + KO_3)/2 - KO_ctrl,
                          levels=KO_design)
KO_fit <- eBayes(KO_fit)
KO_fit2 <- contrasts.fit(KO_fit, KO_cont.mat)
KO_eBay <- eBayes(KO_fit2)
```

Write tables as supplementary data for the records and further interpretation.
```{r}
for(cont in colnames(KO_cont.mat)) {
  cat(paste0("Registering differential expression for ",cont,"\n"),
	file=stdout())
  topTab <- topTable(KO_eBay, coef=cont, number = Inf)
  # DEGs table
  write.table(topTab, file=paste0(DATADIR,cont,"_diffexpr.csv"),sep=",",
                                  row.names=FALSE, col.names=TRUE, quote=FALSE)
#   # Volcano plot
#   plot(KO_eBay$coefficients[,cont], -log10(KO_eBay$p.value[,cont]),
#        xlab="log2-fold-change", ylab="-log10(pvalue)", 
#        cex=0.7, main=cont)
#   # Histogram of p-vals
#   hist(KO_eBay$p.value[,cont], main=cont, xlab="P-values")
}
```

### Over-expression experiment
```{r OE_contrast}
OE_fit <- lmFit(OE_v, OE_design)
OE_cont.mat <- makeContrasts("NKD2_OE"=OE - OE_ctrl,
                          levels=OE_design)
OE_fit <- eBayes(OE_fit)
OE_fit2 <- contrasts.fit(OE_fit, OE_cont.mat)
OE_eBay <- eBayes(OE_fit2)
```

Write tables as supplementary data for the records and further interpretation.
```{r}
for(cont in colnames(OE_cont.mat)) {
  cat(paste0("Registering differential expression for ",cont,"\n"),
	file=stdout())
  topTab <- topTable(OE_eBay, coef=cont, number = Inf)
  # DEGs table
  write.table(topTab, file=paste0(DATADIR,cont,"_diffexpr.csv"),sep=",",
                                  row.names=FALSE, col.names=TRUE, quote=FALSE)
#   # Volcano plot
#   plot(OE_eBay$coefficients[,cont], -log10(OE_eBay$p.value[,cont]),
#        xlab="log2-fold-change", ylab="-log10(pvalue)", 
#        cex=0.7, main=cont)
#   # Histogram of p-vals
#   hist(OE_eBay$p.value[,cont], main=cont, xlab="P-values")
}
```


### Broad statistics on DEGs
Just check how many genes are differentially expressed
```{r}
KO_DGE_cnt <- apply(decideTests(KO_eBay),2, table)
OE_DGE_cnt <- apply(decideTests(OE_eBay),2, table)
DGE_cnt <- cbind(KO_DGE_cnt, OE_DGE_cnt) 
print(DGE_cnt)
```

We could visualize the proportion of differentially expressed genes.
```{r pie_dge, fig.width=7, fig.height=5, dpi=300, warning=FALSE}
NGenes <- colSums(DGE_cnt)

# First dataframe with in/out
DGE_perc <- sweep(DGE_cnt,2, STATS = colSums(DGE_cnt),FUN="/")
DGE_perc <- DGE_perc*100

DGE_perc.df <- data.frame("type"=c(rep("Differential", 3),
				  rep("Unchanged",3)),
			 "value"=c(colSums(DGE_perc[c("-1","1"),]),
				   DGE_perc["0",]),
			 "Contrasts"=rep(colnames(DGE_perc),2))
DGE_perc.df$type <- factor(DGE_perc.df$type, levels=c("Unchanged","Differential"))

# Second dataframe with updown
DGE_perc2 <- sweep(DGE_cnt[c("-1","1"), ],2, STATS = colSums(DGE_cnt[c("-1","1"),]),FUN="/")
DGE_perc2 <- DGE_perc2*100
DGE_perc.df2 <- reshape2::melt(DGE_perc2)
DGE_perc.df2$Var1 <- factor(c("-1"="Downregulated","1"="Upregulated")[as.character(DGE_perc.df2$Var1)],
			   levels=c("Downregulated","Upregulated"))
colnames(DGE_perc.df2) <- c("type","Contrasts","value")

# Pie chart code: https://stackoverflow.com/questions/16184188/ggplot-facet-piechart-placing-text-in-the-middle-of-pie-chart-slices

blank_theme <- theme_minimal()+
  theme(
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

for(cnt in colnames(DGE_cnt)) {
	cat(paste0("Arranging plots for ",cnt,"\n"),file=stdout())
pie1 <- ggplot(subset(DGE_perc.df,Contrasts==cnt), aes(x="", y=value, fill=type)) +
	geom_bar(width=1, stat = "identity", colour="black") +
	coord_polar("y", start=0) +
	scale_fill_manual(values=c("grey","black"), guide = guide_legend(reverse=TRUE)) +
	geom_text(aes(x=1.25,y = cumsum(value)- (0.5*value),
            label = paste0(sprintf("%1.1f",value), "%")), 
		  size=6, col=c("white","black"), family=fontTXT) +
	blank_theme + 
	ylab(paste0("Total ",NGenes[cnt]," genes")) +
	theme(axis.text.x=element_blank(),
	      axis.title.x = element_text(family=fontTXT,size=22),
	      legend.title=element_blank(),
	      legend.text = element_text(size=22, family=fontTXT),
	      legend.position = c(0,0.9),
	      legend.justification = c(0,0)) 

pie2 <- ggplot(subset(DGE_perc.df2,Contrasts==cnt), aes(x="", y=value, fill=type)) +
	geom_bar(width=1, stat = "identity", colour="black") +
	coord_polar("y", start=0) +
	scale_fill_manual(values=c("blue","red"), guide = guide_legend(reverse=TRUE)) +
	geom_text(aes(y = cumsum(value)- (0.5*value), 
            label = paste0(sprintf("%1.1f",value), "%")), 
		  size=6, col=c("black","white"), family=fontTXT) +
	blank_theme +
	ylab(paste0(sprintf("%1.1f",
			    subset(DGE_perc.df, Contrasts==cnt & 
				   type=="Differential")$value),
			    "%",
		    " of the total genes")) +
	theme(axis.text.x=element_blank(),
	      axis.title.x = element_text(family=fontTXT,size=22),
	      legend.title=element_blank(),
	      legend.text = element_text(size=22, family=fontTXT),
	      legend.position = c(0.2,0.9),
	      legend.justification = c(0,0),
	      plot.margin = unit(c(1,1,1,1), "cm"))

grid.arrange(pie1, pie2, nrow=1, top=textGrob(cnt, gp=gpar(fontsize=20)))
}
```

Finally, we show volcano plots and histograms of nominal p-values to understand the
maginude of the differences in the expression.

```{r volcano_cont, fig.width=12, fig.height=4, dpi=300}
dat <- list(NKD2_KO_severe=topTable(KO_eBay, coef="NKD2_KO_severe", number=Inf),
	     NKD2_KO_shallow=topTable(KO_eBay, coef="NKD2_KO_shallow", number=Inf),
	     NKD2_OE=topTable(OE_eBay, coef="NKD2_OE", number=Inf))

for(cnt in names(dat)) dat[[cnt]]$contrast <- cnt
for(cnt in names(dat)) dat[[cnt]]$ID <- rownames(dat[[cnt]])
 
dat <- do.call("rbind", dat)
dat$signif <- ifelse(dat$adj.P.Val < 0.05, "FDR<0.05", "ns")

ggplot(dat, aes(x=logFC, y=-log10(P.Value), colour=signif)) +
	geom_point(alpha=0.4) +
    geom_vline(
        xintercept = c(-1,1),
        col = "red",
        linetype = "dotted",
        size = 1) +
    geom_hline(
        yintercept = c(-log10(0.05)),
        col = "red",
        linetype = "dotted",
        size = 1)+
    scale_color_manual(values=c("orange", "grey")) +
    xlab("log2-fold-change") +
    theme_cowplot() + 
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  axis.text = element_text(family=fontTXT, size=20),
	  ) +
	facet_wrap(~ contrast)
```

```{r hist_cont, fig.width=12, fig.height=4, dpi=300}
ggplot(dat, aes(x=P.Value)) +
	geom_histogram() +
   theme_cowplot() + 
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  axis.text = element_text(family=fontTXT, size=20),
	  ) +
	facet_wrap(~ contrast)

```


## Gene Set Enrichment analysis

First we create a list with the rankings by the moderated-t statistic for each contrast
```{r}
modT <- list(NKD2_KO_severe=KO_eBay$t[,"NKD2_KO_severe"],
	     NKD2_KO_shallow=KO_eBay$t[,"NKD2_KO_shallow"],
	     NKD2_OE=OE_eBay$t[, "NKD2_OE"])
```

### ECM
Herein we explore the impact of NKD2 perturbation in the transcriptome modulation of
matrisome genes using GSEA-preranked with MatrisomeDB.
```{r fgsea_ECM, warning=FALSE}
#	# Alternative option is to use the gene set collection from MSigDB, but we found that
#	# the matrisome DB is more complete in terms of orthologs from mouse.
#	COL <- getGmt("../data/MSigDB/collagen.gmt")
#	COL <- geneIds(COL)

ECM <- read.table("../data/Matrisome/matrisome_hs_masterlist.tsv", sep="\t", 
		  header=TRUE, stringsAsFactors = FALSE)
ECM <- split(ECM$Gene.Symbol,ECM$Category)
ECM <- ECM[names(ECM)!="n/a"]


set.seed(1234)
ECM.res <- lapply(modT, function(rnk) {
  res <- fgsea(pathways = ECM, stats = rnk, eps=0)
  res <- res[order(res$padj, decreasing = FALSE),]
  return(res)
})

for(cont in names(ECM.res)) {
  res.tab <- ECM.res[[cont]]
  res.tab$leadingEdge <- unlist(lapply(res.tab$leadingEdge, function(z) paste(z, collapse = ",")))
  write.table(res.tab, paste0(DATADIR,cont,"_Matrisome_gsea.csv"),sep=",",
              row.names=FALSE, col.names=TRUE, quote=FALSE)
}
# Save it
saveRDS(ECM.res, file=paste0(DATADIR,"/fgsea_ECM.rds"))
```

We make a summary heatmap of this test.
```{r Fig5j, fig.width=5, fig.height=4}
#NOTE: fgsea output is sorted by p-val in the newest version!
ECM.nes <- lapply(ECM.res, function(cnt) setNames(cnt$NES, cnt$pathway))
ECM.nes <- lapply(ECM.nes, function(z) z[order(names(z))])
ECM.nes <- do.call("cbind", ECM.nes)

ECM.padj <- lapply(ECM.res, function(cnt) setNames(cnt$padj, cnt$pathway))
ECM.padj <- lapply(ECM.padj, function(z) z[order(names(z))])
ECM.padj <- do.call("cbind", ECM.padj)
ECM.padj_txt <- pval2txt(ECM.padj)

stopifnot(all(colnames(ECM.nes)==colnames(ECM.padj_txt)))
stopifnot(all(rownames(ECM.nes)==rownames(ECM.padj_txt)))

# Reformat for readibility
colnames(ECM.nes) <- gsub("^NKD2_","",colnames(ECM.nes))
colnames(ECM.padj_txt) <- gsub("^NKD2_","",colnames(ECM.padj_txt))

hp <- makeHP(ECM.nes, ECM.padj_txt)
draw(hp, heatmap_legend_side="bottom")
```

We show the actual collagen genes that are enriched across comparisons, with the
log2-fold-change as magnitude of dysregulation.
```{r hp_KO_collagens, fig.width=4, fig.height=6}
set1 <- ECM.res$NKD2_KO_severe %>% filter(pathway=="Collagens") %>% 
	.$leadingEdge %>% unlist()
set2 <- ECM.res$NKD2_KO_shallow %>% filter(pathway=="Collagens") %>% 
	.$leadingEdge %>% unlist()

genes <- unique(sort(c(set1, set2)))


# Create an empty dummy matrix
mat <- matrix(NA, ncol=2, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat)))
	      )
COL_lfc  <- COL_fdr<- mat

COL_lfc <- KO_eBay$coefficients[genes, ]
COL_fdr <- apply(KO_eBay$p.value, 2, function(pvals) p.adjust(pvals, method="BH")[genes])
TXT <- pval2txt(COL_fdr)

#makeHP(COL_lfc, COL_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(COL_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(COL_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(COL_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
 	show_column_dend=FALSE, show_row_dend=FALSE,	 
# 	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(COL_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", COL_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(COL_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")
```

We extract the leadingEdge genes from each pathway in each contrast from significant hits.
We transform this into a data.frame with membership annotation (pathway).
```{r}
Matrisome_ledge <- lapply(ECM.res,function(z) {
				  res <- z[z$padj < 0.05,c("pathway","leadingEdge")]
				  res2 <- apply(res,1, function(j) data.frame(pathway=rep(j$pathway, 
										  length(j$leadingEdge)),
								      lEdge=unlist(j$leadingEdge)))
				  res3 <- do.call("rbind", res2)
})
Matrisome_ledge <- do.call("rbind", Matrisome_ledge)
rownames(Matrisome_ledge) <- NULL
Matrisome_ledge <- unique(Matrisome_ledge)
head(Matrisome_ledge)

Matrisome_ledge <- Matrisome_ledge[order(Matrisome_ledge$pathway),]
```

Finally, we show their gene expression levels (scaled across samples - by rows) as a heatmap.
```{r hp_bulkNKD2_matrisome, fig.width=8, fig.height=3.5, dpi=300}
fillMatrix <- function(mat, genes) {
	N <- ncol(mat)
	exprGenes <- rownames(mat)
	missing_genes <- setdiff(genes, exprGenes)
	iset_genes <- intersect(genes, exprGenes)
	missing_mat <- matrix(NA, nrow=length(missing_genes), ncol=N,
			      dimnames=list(missing_genes, colnames(mat)))
	
	mat2 <- rbind(mat[iset_genes,], missing_mat)
	mat2 <- mat2[genes, ]
	return(mat2)
}

set.seed(1245) # fix set of random colors in heatmap

KO_mat <- fillMatrix(KO_v$E, Matrisome_ledge$lEdge)
KO_gr <- KO_v$targets[,"group", drop=FALSE]
KO_gr$group <- as.character(KO_gr$group)
KO_gr[which(KO_gr$group=="KO_ctrl"),1] <- "control"
KO_gr[which(KO_gr$group=="KO_1"),1] <- "KO_severe"
KO_gr[which(KO_gr$group=="KO_2"),1] <- "KO_shallow_1"
KO_gr[which(KO_gr$group=="KO_3"),1] <- "KO_shallow_2"
KO_ann <- HeatmapAnnotation(df=KO_gr,
			    col=list(group=c("control"="grey",
					       "KO_severe"="#bf41c6",
					       "KO_shallow_1"="#175f29",
					       "KO_shallow_2"="#82d038")),
			    show_annotation_name = FALSE)

OE_mat <-  fillMatrix(OE_v$E, Matrisome_ledge$lEdge)
OE_gr <- OE_v$targets[,"group", drop=FALSE]
OE_gr$group <- as.character(OE_gr$group)
OE_gr[which(OE_gr$group=="OE_ctrl"),1] <- "control"
OE_gr[which(OE_gr$group=="OE"),1] <- "OverExpr."
OE_ann <- HeatmapAnnotation(df=OE_gr,
			    col=list(group=c("control"="grey",
					     "OverExpr."="#1d4ab7")),
			show_annotation_name = FALSE)


matrisome_df <- data.frame("MatrisomeDB_class"=Matrisome_ledge$pathway,
			   row.names=Matrisome_ledge$ledge)
hr <- rowAnnotation(df=matrisome_df, show_annotation_name=FALSE)

h1 <- hr + Heatmap(t(scale(t(KO_mat))), top_annotation = KO_ann, 
		  heatmap_legend_param=list(legend_direction="horizontal",
					    title_gp=gpar(fontsize=12),
					    labels_gp=gpar(fontsize=12),
					    legend_width=unit(5,"cm")), 
		   show_row_names = FALSE, show_column_names = FALSE,
	      column_title = "NKD2 Knock-out",
	name="Gene Expresion (row scaled)", cluster_rows=FALSE)
h2 <- hr + Heatmap(t(scale(t(OE_mat))), top_annotation = OE_ann, 
		   show_row_names = FALSE, show_column_names = FALSE,
	      column_title = "NKD2 Over-Expression",
	name="Gene Expresion (row scaled)", cluster_rows=FALSE)

h1 + h2
```

We will also visualize the collagen production in this phenotype.
```{r fgsea_KOsevere_ECM, dpi=300, fig.width=8, fig.height=6}
plotGseaTable(ECM, KO_eBay$t[, "NKD2_KO_severe"], 
	      ECM.res[["NKD2_KO_severe"]])  + 
	theme(text=element_text(family=fontTXT, color="black", size=18),
	      axis.text=element_text(family=fontTXT, color="black", size=18)
	      )
```

We finish this section visualizing leadgind edge collagen genes in the volcano plots.
```{r volcano_KOsevere_ECM, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_severe"
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)
path <- "Collagens"
set <- unlist(ECM.res[[cnt]][which(ECM.res[[cnt]]$pathway=="Collagens"), "leadingEdge"])

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

```{r volcano_KOshallow_ECM, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_shallow"
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)
path <- "Collagens"
set <- unlist(ECM.res[[cnt]][which(ECM.res[[cnt]]$pathway=="Collagens"), "leadingEdge"])

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

> Conclusion: NKD2 consistently controls ECM production via the expression of regulators and the expression of collagen genes
themselves in myofibroblasts.

### WNT pathway

NKD2 is a major regulator of WNT pathway. Here we test this pathway using the Hallmark gene set from MSigDB.

```{r fgsea_hallmarks, warning=FALSE}
H <- getGmt("../data/MSigDB/h.all.v7.0.symbols.gmt")
H <- geneIds(H)

set.seed(1234)
H.res <- lapply(modT, function(rnk) {
  res <- fgsea(pathways = H, stats = rnk, eps=0)
  res <- res[order(res$padj, decreasing = FALSE),]
  return(res)
})

for(cont in names(H.res)) {
  res.tab <- H.res[[cont]]
  res.tab$leadingEdge <- unlist(lapply(res.tab$leadingEdge, function(z) paste(z, collapse = ",")))
  write.table(res.tab, paste0(DATADIR,cont,"_Hallmarks_gsea.csv"),sep=",",
              row.names=FALSE, col.names=TRUE, quote=FALSE)
}
```

We report the results for WNT in particular.
```{r}
WNTpath <- "HALLMARK_WNT_BETA_CATENIN_SIGNALING"
lapply(H.res, function(z) z[which(z$pathway==WNTpath), ])
```

We also check the impact of NKD2 perturbation in ligand and receptors from WNT pathway.

```{r hp_WNT_ligrec, fig.width=6, fig.height=5}
KO_genes <- grep("^(WNT|FZD)",rownames(KO_eBay$coef), value=TRUE)
print(KO_genes)

OE_genes <- grep("^(WNT|FZD)",rownames(OE_eBay$coef), value=TRUE)
print(OE_genes)
# Remove a weird gene
KO_genes <- setdiff(KO_genes, "FZD10-DT")
OE_genes <- setdiff(OE_genes, "FZD10-DT")

genes <- unique(sort(c(KO_genes, OE_genes)))


# Create an empty dummy matrix
mat <- matrix(NA, ncol=3, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat), 
				     colnames(OE_cont.mat))
	      		)
	      )
WNT_lfc  <- WNT_fdr<- mat
# Fill out
for(cont in colnames(KO_cont.mat)) WNT_lfc[KO_genes, cont] <- KO_eBay$coef[KO_genes, cont];
for(cont in colnames(OE_cont.mat)) WNT_lfc[OE_genes, cont] <- OE_eBay$coef[OE_genes, cont];

for(cont in colnames(KO_cont.mat)) WNT_fdr[KO_genes, cont] <- p.adjust(KO_eBay$p.value[, cont], method="BH")[KO_genes];
for(cont in colnames(OE_cont.mat)) WNT_fdr[OE_genes, cont] <- p.adjust(OE_eBay$p.value[, cont], method="BH")[OE_genes];


TXT <- pval2txt(WNT_fdr)

#makeHP(WNT_lfc, WNT_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(WNT_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(WNT_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(WNT_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
# 	show_column_dend=FALSE, show_row_dend=FALSE,	 
	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(WNT_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", WNT_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(WNT_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")
```

> Conclusion: WNT dysregulation is not conclusive in the experiment. GSEA shows a common trend towards enrichment of WNT
in the control group. That means, it was enriched in control phenotype in KO-shallow towards controls (FDR < 0.05, down-regulation).
Following same trend in KO-severe, rather nonsignificant though (FDR>0.05). But also it was the case for NKD2 over-expression (marginal FDR=0.052), 
which is an opposite perturbation to KO, and it would be expected to be enriched in opposite direction. 
On the other hand, we observe that ligand-receptor of WNT was significantlly dysregulated upon NKD2 perturbation, mainly in the 
KO severe phenotype.


### Single-cell derived markers
Similarly to previous section, we also checked the markers derived from single-cell that are
characteristic of myofibroblasts, and its transcriptome modulation upon NKD2 perturbation.

```{r hp_scmarkers, fig.width=6, fig.height=5}
genes <- scan("../data/bulk-rnaseq/singlecell_list.txt", what="character")

KO_genes <- intersect(genes, rownames(KO_v))
OE_genes <- intersect(genes, rownames(OE_v))

# Create an empty dummy matrix
mat <- matrix(NA, ncol=3, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat), 
				     colnames(OE_cont.mat))
	      		)
	      )
SC_lfc  <- SC_fdr<- mat
# Fill out
for(cont in colnames(KO_cont.mat)) SC_lfc[KO_genes, cont] <- KO_eBay$coef[KO_genes, cont];
for(cont in colnames(OE_cont.mat)) SC_lfc[OE_genes, cont] <- OE_eBay$coef[OE_genes, cont];

for(cont in colnames(KO_cont.mat)) SC_fdr[KO_genes, cont] <- p.adjust(KO_eBay$p.value[, cont], method="BH")[KO_genes];
for(cont in colnames(OE_cont.mat)) SC_fdr[OE_genes, cont] <- p.adjust(OE_eBay$p.value[, cont], method="BH")[OE_genes];


TXT <- pval2txt(SC_fdr)

#makeHP(SC_lfc, SC_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(SC_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(SC_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(SC_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
# 	show_column_dend=FALSE, show_row_dend=FALSE,	 
	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(SC_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", SC_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(SC_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")
```

> Markers for internal use. Actually it is found Collagen genes dysregulated as previous section. 
But also FOS expression.

## Save processed data
We are going to save the eBayes object for downstream analysis.
```{r}
# The Elist object, with voom-transformed gene expression
saveRDS(KO_v, paste0(DATADIR,"/KO_v.rds"))
saveRDS(OE_v, paste0(DATADIR,"/OE_v.rds"))
# The eBayes object from the two contrasts
saveRDS(KO_eBay, paste0(DATADIR,"/KO_eBay.rds"))
saveRDS(OE_eBay, paste0(DATADIR,"/OE_eBay.rds"))
```

## Session info

```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```

