---
title: "Differential expression analysis in Myofibroblast with NKD2 perturbations"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

> Rafael kramann's lab has established one cell line from myofibroblasts from kidney bipsies with CKD.
Then this cell line was transformed with NKD-KO, which generated three different clones carrying NKD2-KO.
In addition, myofibroblasts were also transformed with NKD2- over-expression. These gene perturbations are
compared with their corresponding empty-vector controls. 

Previous script ([00_EDA.md](./00_EDA.md)) shows that:

* Biological replicates cluster together.

* Theree is one KO clone that present much more differences to the other KO clones. 
  With experimental observations,   that clone is actually a clone with a fibrotic 
  severe phenotype.

## Set-up environment
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Define random seed for reproduciblity, and output folder.
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./01_DGE_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png', 'pdf'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);

# If already exists, clean dirs?
clean_dirs <- TRUE
if(clean_dirs) {
	unlink(list.files(OUTDIR, full.names=TRUE, recursive = TRUE))
}
```


## Load libraries
Essential R libraries for the analysis.
```{r}
library(limma)
library(edgeR)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(GSEABase)
library(fgsea)
library(reshape2)
library(gridExtra)
library(ComplexHeatmap)
library(grid)
fontTXT <- "sans"
```
## Load data and normalize

```{r}
### 1 Load data
# Source: https://bioconductor.org/packages/release/bioc/vignettes/tximport/inst/doc/tximport.html#limma-voom
load("../data/bulk-rnaseq/forJavier.RData")
# The sample IDs are stored in the folder from the file that was imported to R. We recall them
sname <- sapply(dirname(filenames),function(z) paste(strsplit(z, split="_")[[1]][1:2],collapse="_"))
# We get the read counts per sample
cnt <- txi$counts
# Rename samples
colnames(cnt) <- sname
# Get the biological conditions
group_str <- as.character(group)
group_str <- gsub("-","_", group_str)
group_str <- gsub("OverExpr","OE",group_str)
```

### Knock-out experiment
Define data experiment
```{r}
KO_idx <- grep("^KO", group_str)
KO_cnt <- cnt[, KO_idx]
KO_gr <- factor(group_str[KO_idx])
KO_gr <- relevel(KO_gr, ref="KO_ctrl")
```

```{r}
## 2 Create DGElist (edgeR package, followed by TMM default normalization method)
KO_y <- DGEList(counts = KO_cnt, group = KO_gr, genes = rownames(KO_cnt))
# Filtering
KO_keep <- filterByExpr(KO_y)
table(KO_keep)

# Apply filtering
KO_y <- KO_y[KO_keep, ]
KO_y <- calcNormFactors(KO_y)

## 3 Create experimental design
stopifnot(all(KO_gr == KO_y$samples$group))
KO_design <- model.matrix(~ 0 + KO_gr) # wo/ intercept
colnames(KO_design) <- gsub("^KO_gr", "", colnames(KO_design))
KO_v <- voom(KO_y, KO_design)

# Retrieve only gene symbols for downstream analysis
rownames(KO_v) <- sapply(rownames(KO_v), function(z) strsplit(z, split=";")[[1]][2])
```

### Over-expression experiment
Define data experiment
```{r}
OE_idx <- grep("^OE", group_str)
OE_cnt <- cnt[, OE_idx]
OE_gr <- factor(group_str[OE_idx])
OE_gr <- relevel(OE_gr, ref="OE_ctrl")
```

```{r}
## 2 Create DGElist (edgeR package, followed by TMM default normalization method)
OE_y <- DGEList(counts = OE_cnt, group = OE_gr, genes = rownames(OE_cnt))
# Filtering
OE_keep <- filterByExpr(OE_y)
table(OE_keep)

# Apply filtering
OE_y <- OE_y[OE_keep, ]
OE_y <- calcNormFactors(OE_y)

## 3 Create experimental design
stopifnot(all(OE_gr == OE_y$samples$group))
OE_design <- model.matrix(~ 0 + OE_gr) # wo/ intercept
colnames(OE_design) <- gsub("^OE_gr", "", colnames(OE_design))
OE_v <- voom(OE_y, OE_design)

# Retrieve only gene symbols for downstream analysis
rownames(OE_v) <- sapply(rownames(OE_v), function(z) strsplit(z, split=";")[[1]][2])
```

## Test for differential expression

### Knock-out experiment
```{r KO_contrast}
KO_fit <- lmFit(KO_v, KO_design)
KO_cont.mat <- makeContrasts("NKD2_KO_severe"=KO_1 - KO_ctrl,
			  "NKD2_KO_shallow"=(KO_2 + KO_3)/2 - KO_ctrl,
                          levels=KO_design)
KO_fit <- eBayes(KO_fit)
KO_fit2 <- contrasts.fit(KO_fit, KO_cont.mat)
KO_eBay <- eBayes(KO_fit2)
```

Write tables as supplementary data for the records and further interpretation.
```{r}
for(cont in colnames(KO_cont.mat)) {
  cat(paste0("Registering differential expression for ",cont,"\n"),
	file=stdout())
  topTab <- topTable(KO_eBay, coef=cont, number = Inf)
  # DEGs table
  write.table(topTab, file=paste0(DATADIR,cont,"_diffexpr.csv"),sep=",",
                                  row.names=FALSE, col.names=TRUE, quote=FALSE)
  # Volcano plot
  plot(KO_eBay$coefficients[,cont], -log10(KO_eBay$p.value[,cont]),
       xlab="log2-fold-change", ylab="-log10(pvalue)", 
       cex=0.7, main=cont)
  # Histogram of p-vals
  hist(KO_eBay$p.value[,cont], main=cont, xlab="P-values")
}
```

### Knock-out experiment
```{r OE_contrast}
OE_fit <- lmFit(OE_v, OE_design)
OE_cont.mat <- makeContrasts("NKD2_OE"=OE - OE_ctrl,
                          levels=OE_design)
OE_fit <- eBayes(OE_fit)
OE_fit2 <- contrasts.fit(OE_fit, OE_cont.mat)
OE_eBay <- eBayes(OE_fit2)
```

Write tables as supplementary data for the records and further interpretation.
```{r}
for(cont in colnames(KO_cont.mat)) {
  cat(paste0("Registering differential expression for ",cont,"\n"),
	file=stdout())
  topTab <- topTable(KO_eBay, coef=cont, number = Inf)
  # DEGs table
  write.table(topTab, file=paste0(DATADIR,cont,"_diffexpr.csv"),sep=",",
                                  row.names=FALSE, col.names=TRUE, quote=FALSE)
  # Volcano plot
  plot(KO_eBay$coefficients[,cont], -log10(KO_eBay$p.value[,cont]),
       xlab="log2-fold-change", ylab="-log10(pvalue)", 
       cex=0.7, main=cont)
  # Histogram of p-vals
  hist(KO_eBay$p.value[,cont], main=cont, xlab="P-values")
}
```


## Broad statistics on DEGs
Just check how many genes are differentially expressed
```{r}
KO_DGE_cnt <- apply(decideTests(KO_eBay),2, table)
OE_DGE_cnt <- apply(decideTests(OE_eBay),2, table)
DGE_cnt <- cbind(KO_DGE_cnt, OE_DGE_cnt) 
print(DGE_cnt)
```

We could visualize the proportion of differentially expressed genes.
```{r pie_dge, fig.width=7, fig.height=5, dpi=300, warning=FALSE}
NGenes <- colSums(DGE_cnt)

# First dataframe with in/out
DGE_perc <- sweep(DGE_cnt,2, STATS = colSums(DGE_cnt),FUN="/")
DGE_perc <- DGE_perc*100

DGE_perc.df <- data.frame("type"=c(rep("Differential", 3),
				  rep("Unchanged",3)),
			 "value"=c(colSums(DGE_perc[c("-1","1"),]),
				   DGE_perc["0",]),
			 "Contrasts"=rep(colnames(DGE_perc),2))
DGE_perc.df$type <- factor(DGE_perc.df$type, levels=c("Unchanged","Differential"))

# Second dataframe with updown
DGE_perc2 <- sweep(DGE_cnt[c("-1","1"), ],2, STATS = colSums(DGE_cnt[c("-1","1"),]),FUN="/")
DGE_perc2 <- DGE_perc2*100
DGE_perc.df2 <- reshape2::melt(DGE_perc2)
DGE_perc.df2$Var1 <- factor(c("-1"="Downregulated","1"="Upregulated")[as.character(DGE_perc.df2$Var1)],
			   levels=c("Downregulated","Upregulated"))
colnames(DGE_perc.df2) <- c("type","Contrasts","value")

# Pie chart code: https://stackoverflow.com/questions/16184188/ggplot-facet-piechart-placing-text-in-the-middle-of-pie-chart-slices

blank_theme <- theme_minimal()+
  theme(
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

for(cnt in colnames(DGE_cnt)) {
	cat(paste0("Arranging plots for ",cnt,"\n"),file=stdout())
pie1 <- ggplot(subset(DGE_perc.df,Contrasts==cnt), aes(x="", y=value, fill=type)) +
	geom_bar(width=1, stat = "identity", colour="black") +
	coord_polar("y", start=0) +
	scale_fill_manual(values=c("grey","black"), guide = guide_legend(reverse=TRUE)) +
	geom_text(aes(x=1.25,y = cumsum(value)- (0.5*value),
            label = paste0(sprintf("%1.1f",value), "%")), 
		  size=6, col=c("white","black"), family=fontTXT) +
	blank_theme + 
	ylab(paste0("Total ",NGenes[cnt]," genes")) +
	theme(axis.text.x=element_blank(),
	      axis.title.x = element_text(family=fontTXT,size=22),
	      legend.title=element_blank(),
	      legend.text = element_text(size=22, family=fontTXT),
	      legend.position = c(0,0.9),
	      legend.justification = c(0,0)) 

pie2 <- ggplot(subset(DGE_perc.df2,Contrasts==cnt), aes(x="", y=value, fill=type)) +
	geom_bar(width=1, stat = "identity", colour="black") +
	coord_polar("y", start=0) +
	scale_fill_manual(values=c("blue","red"), guide = guide_legend(reverse=TRUE)) +
	geom_text(aes(y = cumsum(value)- (0.5*value), 
            label = paste0(sprintf("%1.1f",value), "%")), 
		  size=6, col=c("black","white"), family=fontTXT) +
	blank_theme +
	ylab(paste0(sprintf("%1.1f",
			    subset(DGE_perc.df, Contrasts==cnt & 
				   type=="Differential")$value),
			    "%",
		    " of the total genes")) +
	theme(axis.text.x=element_blank(),
	      axis.title.x = element_text(family=fontTXT,size=22),
	      legend.title=element_blank(),
	      legend.text = element_text(size=22, family=fontTXT),
	      legend.position = c(0.2,0.9),
	      legend.justification = c(0,0),
	      plot.margin = unit(c(1,1,1,1), "cm"))

grid.arrange(pie1, pie2, nrow=1, top=textGrob(cnt, gp=gpar(fontsize=20)))
}
```

## WNT lig-rec pathway
For the records, we check the impact in genes encoding ligand and receptors from the
WNT pathway upon NKD2 perturbation.

```{r pval2txt}
pval2txt <- function(padj) {
padj_txt <- apply(padj, c(1,2), function(pval) {
	if(!is.na(pval)) {
		   if(pval< 0.001) {
			   txt <- "***"
		   } else if (pval < 0.01) {
			   txt <- "**"
		   } else if (pval < 0.05) {
			   txt <- "*"
		   } else {
			   txt <- "ns"
		   } 
	} else {
		txt <- "n/a"
	}
	return(txt)
})

return(padj_txt)
}
```

```{r hp_WNT_ligrec, fig.width=6, fig.height=5}
KO_genes <- grep("^(WNT|FZD)",rownames(KO_eBay$coef), value=TRUE)
print(KO_genes)

OE_genes <- grep("^(WNT|FZD)",rownames(OE_eBay$coef), value=TRUE)
print(OE_genes)
# Remove a weird gene
KO_genes <- setdiff(KO_genes, "FZD10-DT")
OE_genes <- setdiff(OE_genes, "FZD10-DT")

genes <- unique(sort(c(KO_genes, OE_genes)))


# Create an empty dummy matrix
mat <- matrix(NA, ncol=3, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat), 
				     colnames(OE_cont.mat))
	      		)
	      )
WNT_lfc  <- WNT_fdr<- mat
# Fill out
for(cont in colnames(KO_cont.mat)) WNT_lfc[KO_genes, cont] <- KO_eBay$coef[KO_genes, cont];
for(cont in colnames(OE_cont.mat)) WNT_lfc[OE_genes, cont] <- OE_eBay$coef[OE_genes, cont];

for(cont in colnames(KO_cont.mat)) WNT_fdr[KO_genes, cont] <- p.adjust(KO_eBay$p.value[, cont], method="BH")[KO_genes];
for(cont in colnames(OE_cont.mat)) WNT_fdr[OE_genes, cont] <- p.adjust(OE_eBay$p.value[, cont], method="BH")[OE_genes];


TXT <- pval2txt(WNT_fdr)

#makeHP(WNT_lfc, WNT_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(WNT_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(WNT_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(WNT_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
# 	show_column_dend=FALSE, show_row_dend=FALSE,	 
	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(WNT_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", WNT_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(WNT_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")
```

## Single-cell derived markers
Similarly to previous section, we also checked the markers derived from single-cell that are
characteristic of myofibroblasts, and its transcriptome modulation upon NKD2 perturbation.

```{r hp_scmarkers, fig.width=6, fig.height=5}
genes <- scan("../data/bulk-rnaseq/singlecell_list.txt", what="character")

KO_genes <- intersect(genes, rownames(KO_v))
OE_genes <- intersect(genes, rownames(OE_v))

# Create an empty dummy matrix
mat <- matrix(NA, ncol=3, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat), 
				     colnames(OE_cont.mat))
	      		)
	      )
SC_lfc  <- SC_fdr<- mat
# Fill out
for(cont in colnames(KO_cont.mat)) SC_lfc[KO_genes, cont] <- KO_eBay$coef[KO_genes, cont];
for(cont in colnames(OE_cont.mat)) SC_lfc[OE_genes, cont] <- OE_eBay$coef[OE_genes, cont];

for(cont in colnames(KO_cont.mat)) SC_fdr[KO_genes, cont] <- p.adjust(KO_eBay$p.value[, cont], method="BH")[KO_genes];
for(cont in colnames(OE_cont.mat)) SC_fdr[OE_genes, cont] <- p.adjust(OE_eBay$p.value[, cont], method="BH")[OE_genes];


TXT <- pval2txt(SC_fdr)

#makeHP(SC_lfc, SC_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(SC_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(SC_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(SC_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
# 	show_column_dend=FALSE, show_row_dend=FALSE,	 
	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(SC_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", SC_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(SC_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")

```


## GSEA

Some handy functions for visualization
Make a heatmap with text showing significant hits.
```{r make_hp}
### Make a heatmap of the GSEA results for the 3 contrasts
# NES is a matrix of NES
# PADJ is a matrix of strings for each cell
makeHP <- function(NES, TXT) {

hp <- Heatmap(NES, column_names_side="top", row_names_side="left",
	      name="Normalized Enrichment Score (NES)",
	      heatmap_legend_param=list(direction="horizontal",
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
	show_column_dend=FALSE, show_row_dend=FALSE,	 
 	column_names_rot=0,
	row_names_gp=gpar(fontsize=12),
    	row_names_max_width = max_text_width(
        			rownames(NES), 
        			gp = gpar(fontsize = 12)),
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		grid.text(TXT[i, j], x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(NES[i,j] < -2,
						 "white","black"))
			  )
	}
)
return(hp)
}
```

Barplot with significantly enriched pathways in a phenotype.
```{r}
makeBP <- function(fgsea_out) {

	df <- as.data.frame(fgsea_out[,c("pathway","NES","padj")])
	df <- df[order((df$NES), decreasing=FALSE),]
	df$pathway <- gsub("^HALLMARK","",df$pathway)
	df$pathway <- gsub("_"," ",df$pathway)
	df$pathway <- factor(df$pathway, levels=unique(df$pathway))

gg <- ggplot(subset(df, padj < 0.05), aes(x=pathway, y=abs(NES), fill=NES)) + 
    geom_bar(stat="identity", width=0.5) +
    scale_fill_gradient2(low="blue",mid="white", high="red") + 
    theme_cowplot() +
    coord_flip() +
    guides(fill = guide_colourbar(title="Normalized Enrichment Score (NES)",
                   title.position="top"
                   )

           ) +
    ylab("Absolute (NES)") +
    theme(legend.position="bottom",
          legend.key.width=unit(1.5,"cm"),
          legend.title=element_text(family=fontTXT, size=16),
          legend.text=element_text(family=fontTXT, size=16),
          axis.title.y=element_blank(),
          axis.text.y=element_text(family=fontTXT, size=14),
          axis.text.x=element_text(family=fontTXT, size=18),
          axis.title.x=element_text(family=fontTXT, size=18))

    return(gg)
}
```


### HALLMARK collection
We conclude this section with the charaterization of the cellular state in the two experiments 
using the hallmark gene set collection from MsigDB.
We are particularly interested in PHD2cKO Macrophages. 

First we create a list with the rankings by the moderated-t statistic for each contrast
```{r}
modT <- list(NKD2_KO_severe=KO_eBay$t[,"NKD2_KO_severe"],
	     NKD2_KO_shallow=KO_eBay$t[,"NKD2_KO_shallow"],
	     NKD2_OE=OE_eBay$t[, "NKD2_OE"])
```

```{r fgsea_hallmarks, warning=FALSE}
H <- getGmt("../data/MSigDB/h.all.v7.0.symbols.gmt")
H <- geneIds(H)

set.seed(1234)
H.res <- lapply(modT, function(rnk) {
  res <- fgsea(pathways = H, stats = rnk, eps=0)
  res <- res[order(res$padj, decreasing = FALSE),]
  return(res)
})

for(cont in names(H.res)) {
  res.tab <- H.res[[cont]]
  res.tab$leadingEdge <- unlist(lapply(res.tab$leadingEdge, function(z) paste(z, collapse = ",")))
  write.table(res.tab, paste0(DATADIR,cont,"_Hallmarks_gsea.csv"),sep=",",
              row.names=FALSE, col.names=TRUE, quote=FALSE)
}
```

We report the significant (FDR<0.05) hallmarks enriched in the KO severe phenotype.

```{r barplot_GSEA_hallmarks_KOsevere, fig.width=8, fig.height=8, dpi=300, warning=FALSE}
makeBP(H.res$NKD2_KO_severe) + ggtitle("NKD2_KO_severe")
```
Similarly for the KO shallow phenotype,
```{r barplot_GSEA_hallmarks_KOshallow, fig.width=8, fig.height=8, dpi=300, warning=FALSE}
makeBP(H.res$NKD2_KO_shallow) + ggtitle("NKD2_KO_shallow")
```

And for the OE experiment,
```{r barplot_GSEA_hallmarks_OE, fig.width=8, fig.height=8, dpi=300, warning=FALSE}
makeBP(H.res$NKD2_OE) + ggtitle("NKD2_KO_OE")
```

We could summarize the plots above as a heatmap
```{r hp_GSEA_hallmarks, fig.width=7, fig.height=10}
#NOTE: fgsea output is sorted by p-val in the newest version!
H.nes <- lapply(H.res, function(cnt) setNames(cnt$NES, cnt$pathway))
H.nes <- lapply(H.nes, function(z) z[order(names(z))])
H.nes <- do.call("cbind", H.nes)

H.padj <- lapply(H.res, function(cnt) setNames(cnt$padj, cnt$pathway))
H.padj <- lapply(H.padj, function(z) z[order(names(z))])
H.padj <- do.call("cbind", H.padj)
H.padj_txt <- pval2txt(H.padj)

stopifnot(all(colnames(H.nes)==colnames(H.padj_txt)))
stopifnot(all(rownames(H.nes)==rownames(H.padj_txt)))

# Reformat for readibility
rownames(H.nes) <- gsub("^HALLMARK_","",rownames(H.nes))
rownames(H.padj_txt) <- gsub("^HALLMARK_","",rownames(H.padj_txt))
colnames(H.nes) <- gsub("^NKD2_","",colnames(H.nes))
colnames(H.padj_txt) <- gsub("^NKD2","",colnames(H.padj_txt))

hp <- makeHP(H.nes, H.padj_txt)
draw(hp, heatmap_legend_side="bottom")
```

### ECM
Even more interesting, it is the impact of NKD2 perturbation in the transcriptome modulation of
matrisome genes.
```{r fgsea_ECM, warning=FALSE}
#	# Alternative option is to use the gene set collection from MSigDB, but we found that
#	# the matrisome DB is more complete in terms of orthologs from mouse.
#	COL <- getGmt("../data/MSigDB/collagen.gmt")
#	COL <- geneIds(COL)

ECM <- read.table("../data/Matrisome/matrisome_hs_masterlist.tsv", sep="\t", 
		  header=TRUE, stringsAsFactors = FALSE)
ECM <- split(ECM$Gene.Symbol,ECM$Category)
ECM <- ECM[names(ECM)!="n/a"]


set.seed(1234)
ECM.res <- lapply(modT, function(rnk) {
  res <- fgsea(pathways = ECM, stats = rnk, eps=0)
  res <- res[order(res$padj, decreasing = FALSE),]
  return(res)
})

for(cont in names(ECM.res)) {
  res.tab <- ECM.res[[cont]]
  res.tab$leadingEdge <- unlist(lapply(res.tab$leadingEdge, function(z) paste(z, collapse = ",")))
  write.table(res.tab, paste0(DATADIR,cont,"_Matrisome_gsea.csv"),sep=",",
              row.names=FALSE, col.names=TRUE, quote=FALSE)
}
# Save it
saveRDS(ECM.res, file=paste0(DATADIR,"/fgsea_ECM.rds"))
```

We make a summary heatmap of this test.
```{r hp_GSEA_Matrisome, fig.width=5, fig.height=4}
#NOTE: fgsea output is sorted by p-val in the newest version!
ECM.nes <- lapply(ECM.res, function(cnt) setNames(cnt$NES, cnt$pathway))
ECM.nes <- lapply(ECM.nes, function(z) z[order(names(z))])
ECM.nes <- do.call("cbind", ECM.nes)

ECM.padj <- lapply(ECM.res, function(cnt) setNames(cnt$padj, cnt$pathway))
ECM.padj <- lapply(ECM.padj, function(z) z[order(names(z))])
ECM.padj <- do.call("cbind", ECM.padj)
ECM.padj_txt <- pval2txt(ECM.padj)

stopifnot(all(colnames(ECM.nes)==colnames(ECM.padj_txt)))
stopifnot(all(rownames(ECM.nes)==rownames(ECM.padj_txt)))

# Reformat for readibility
colnames(ECM.nes) <- gsub("^NKD2_","",colnames(ECM.nes))
colnames(ECM.padj_txt) <- gsub("^NKD2_","",colnames(ECM.padj_txt))

hp <- makeHP(ECM.nes, ECM.padj_txt)
draw(hp, heatmap_legend_side="bottom")
```

```{r hp_KO_collagens, fig.width=4, fig.height=6}
set1 <- unlist(ECM.res[[cnt]][which(ECM.res[["NKD2_KO_severe"]]$pathway=="Collagens"), 
	       "leadingEdge"])
set2 <- unlist(ECM.res[[cnt]][which(ECM.res[["NKD2_KO_shallow"]]$pathway=="Collagens"), 
	       "leadingEdge"])

genes <- unique(sort(c(set1, set2)))


# Create an empty dummy matrix
mat <- matrix(NA, ncol=2, nrow=length(genes),
	      dimnames=list(genes, c(colnames(KO_cont.mat)))
	      )
COL_lfc  <- COL_fdr<- mat

COL_lfc <- KO_eBay$coefficients[genes, ]
COL_fdr <- apply(KO_eBay$p.value, 2, function(pvals) p.adjust(pvals, method="BH")[genes])
TXT <- pval2txt(COL_fdr)

#makeHP(COL_lfc, COL_fdr_txt)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(COL_lfc), na.rm=TRUE)),
		 -2,0,2,
		 ceiling(max(as.vector(COL_lfc), na.rm=TRUE)))
legend_labs <- unique(legend_labs)
hp <- Heatmap(COL_lfc, column_names_side="top", row_names_side="left",
	      name="log2-fold-change",col=col_fun,
	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
 	show_column_dend=FALSE, show_row_dend=FALSE,	 
# 	      cluster_rows = FALSE, cluster_columns=FALSE,
 	column_names_rot=0,
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		if(!is.na(COL_lfc[i,j])) {
		grid.text(paste0(sprintf("%.2f", COL_lfc[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(COL_lfc[i,j] < -1,
						 "white","black"))
			  )
		}
	}
)

draw(hp, heatmap_legend_side="bottom")
```


## Visualization of differentially expressed genes
We will prompt with volcano plots showing leading edge genes from each interesting pathway
that we found in the functional analysis.
```{r make_volcano}
make_volcano <- function(topTab, set) {

if(any(!set %in% topTab$ID)) {
	cat(paste("The following genes are not present in the gene expr matrix:","\n",
		  paste(setdiff(set, rownames(topTab)), collapse=","),
		  "\n"),
		  file=stdout())
	set <- intersect(set, rownames(topTab))
}

topTab$significant <- ifelse(topTab$adj.P.Val < 0.05,"FDR<0.05","n.s.")
topTab$show <- topTab$ID %in% set
 
lfc <- topTab$logFC
mxpval <- max(-log10(topTab$P.Value))

p = ggplot(topTab, aes(logFC, -log10(P.Value))) +
    geom_point(aes(col=significant)) +
    scale_color_manual(values=c("red", "grey"))
  
  
p <- p+geom_label_repel(data=topTab[topTab$show,],
                          aes(label=ID), family=fontTXT, size=4,
                          force=2,
			  max.iter=1e4,
# 			  xlim = c(ceiling(max(lfc))+1, ceiling(max(lfc))+2),
 			  xlim = c(floor(min(lfc))*1.5, floor(min(lfc)*1.2)),
# 			  xlim = c(min(lfc), 0),
                           nudge_y=0.05, 
			  direction = "y", 
			  segment.size = 0.3) + 
    coord_cartesian(xlim = c(floor(min(lfc)), 
			     ceiling(max(lfc))), 
		    clip = "off") +
    scale_y_continuous(position="right") + 
    theme_cowplot() +
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  legend.justification=c(1,0),
	  axis.text = element_text(family=fontTXT, size=20),
     	plot.margin = unit(c(0.3, 2.6, 1.0, 4.0), "cm") 
	  )
return(p)

}
```
### NKD2-KO severe phenotype

```{r volcano_KOsevere_WNT, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_severe"
path <- "HALLMARK_WNT_BETA_CATENIN_SIGNALING"
set <- unlist(H.res[[cnt]][which(H.res[[cnt]]$pathway==path),"leadingEdge"])
# Mac_interesting_genes <- H[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]] 
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

```{r volcano_KOsevere_TGFb, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_severe"
path <- "HALLMARK_TGF_BETA_SIGNALING"
set <- unlist(H.res[[cnt]][which(H.res[[cnt]]$pathway==path),"leadingEdge"])
# Mac_interesting_genes <- H[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]] 
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```



We will also visualize the collagen production in this phenotype
```{r fgsea_KOsevere_ECM, dpi=300, fig.width=8, fig.height=6}
# plotEnrichment(ECM[["Collagens"]], rnk) + labs(title="Fibroblasts PHD2")
plotGseaTable(ECM, KO_eBay$t[, "NKD2_KO_severe"], 
	      ECM.res[["NKD2_KO_severe"]])  + 
	theme(text=element_text(family=fontTXT, color="black", size=18),
	      axis.text=element_text(family=fontTXT, color="black", size=18)
	      )
```

```{r volcano_KOsevere_ECM, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_severe"
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)
path <- "Collagens"
set <- unlist(ECM.res[[cnt]][which(ECM.res[[cnt]]$pathway=="Collagens"), "leadingEdge"])

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

```{r volcano_KOshallow_ECM, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_shallow"
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)
path <- "Collagens"
set <- unlist(ECM.res[[cnt]][which(ECM.res[[cnt]]$pathway=="Collagens"), "leadingEdge"])

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```


### NKD2-KO shallow phenotype
```{r volcano_KOshallow_WNT, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_shallow"
path <- "HALLMARK_WNT_BETA_CATENIN_SIGNALING"
set <- unlist(H.res[[cnt]][which(H.res[[cnt]]$pathway==path),"leadingEdge"])
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

```{r volcano_KOshallow_TGFb, dpi=300, fig.width=8, fig.height=7, warning=FALSE}
cnt <- "NKD2_KO_shallow"
path <- "HALLMARK_TGF_BETA_SIGNALING"
set <- unlist(H.res[[cnt]][which(H.res[[cnt]]$pathway==path),"leadingEdge"])
# Mac_interesting_genes <- H[["HALLMARK_WNT_BETA_CATENIN_SIGNALING"]] 
topTab <- topTable(KO_eBay, coef=cnt, number = Inf)

(make_volcano(topTab, set) + labs(title=cnt, subtitle = path))
```

```{r volcano_cont, fig.width=12, fig.height=4, dpi=300}
dat <- list(NKD2_KO_severe=topTable(KO_eBay, coef="NKD2_KO_severe", number=Inf),
	     NKD2_KO_shallow=topTable(KO_eBay, coef="NKD2_KO_shallow", number=Inf),
	     NKD2_OE=topTable(OE_eBay, coef="NKD2_OE", number=Inf))

for(cnt in names(dat)) dat[[cnt]]$contrast <- cnt
for(cnt in names(dat)) dat[[cnt]]$ID <- rownames(dat[[cnt]])
# for(cnt in names(dat)) {
# 	dat[[cnt]]$show <- FALSE
# 	dat[[cnt]]$show[1:20] <- TRUE
# }
# 
dat <- do.call("rbind", dat)
dat$signif <- ifelse(dat$adj.P.Val < 0.05, "FDR<0.05", "ns")

ggplot(dat, aes(x=logFC, y=-log10(P.Value), colour=signif)) +
	geom_point(alpha=0.4) +
    geom_vline(
        xintercept = c(-1,1),
        col = "red",
        linetype = "dotted",
        size = 1) +
    geom_hline(
        yintercept = c(-log10(0.05)),
        col = "red",
        linetype = "dotted",
        size = 1)+
    scale_color_manual(values=c("orange", "grey")) +
    xlab("log2-fold-change") +
    theme_cowplot() + 
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  axis.text = element_text(family=fontTXT, size=20),
	  ) +
	facet_wrap(~ contrast)
```

```{r hist_cont, fig.width=12, fig.height=4, dpi=300}
ggplot(dat, aes(x=P.Value)) +
	geom_histogram() +
   theme_cowplot() + 
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  legend.position = "bottom",
	  axis.text = element_text(family=fontTXT, size=20),
	  ) +
	facet_wrap(~ contrast)

```

## Save processed data
We are going to save the eBayes object for downstream analysis.
```{r}
# The Elist object, with voom-transformed gene expression
saveRDS(KO_v, paste0(DATADIR,"/KO_v.rds"))
saveRDS(OE_v, paste0(DATADIR,"/OE_v.rds"))
# The eBayes object from the two contrasts
saveRDS(KO_eBay, paste0(DATADIR,"/KO_eBay.rds"))
saveRDS(OE_eBay, paste0(DATADIR,"/OE_eBay.rds"))
```

## Session info

```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```

