---
title: "Inference of pathway activities in Myelofibroblast with NKD2 gene perturbations in kidney"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

Herein we explore pathway activities on the two experiments using PROGENy. We perform two inferences in 
parallel for each experiment, using single-sample PROGENy estimates and gene-wise permutation PROGENy.

## Set-up environment
Define randome seed and output folder.
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./02_pathway_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png', 'pdf'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
# If already exists, clean dirs?
clean_dirs <- TRUE
if(clean_dirs) {
	unlink(list.files(OUTDIR, full.names=TRUE, recursive = TRUE))
}
```

## Load libraries
Essential R libraries.
```{r}
library(limma)
library(progeny)
library(broom)
library(magrittr)
library(ggplot2)
suppressPackageStartupMessages(library(ComplexHeatmap))
```

## Load data
We use the output data from previous step (01). In particular,
the eBay object from limma which contains the contrasts and statistics.
```{r}
KO_v <- readRDS("./01_DGE_output/data/KO_v.rds")
KO_eBay <- readRDS("./01_DGE_output/data/KO_eBay.rds")

OE_v <- readRDS("./01_DGE_output/data/OE_v.rds")
OE_eBay <- readRDS("./01_DGE_output/data/OE_eBay.rds")
```

## PROGENy analysis

### PROGENy single-sample
First, we define a function to pathway-wise test differences using a linear model.
```{r}
test_lm <- function(pathways, controls) {
	result <- apply(pathways, 1, function(y) {
				broom::tidy(lm(y ~ !controls)) %>%
				dplyr::filter(term == "!controlsTRUE") %>%
				dplyr::select(-term)
	})
	res <- dplyr::mutate(dplyr::bind_rows(result), 
			     pathway=names(result))
	return(res)
}
```

We are going to calculate progeny scores without scaling the data. 
We are going to scale it with respect the control conditions in each contrast.
```{r KO_ssprog}
KO_prog <- progeny(KO_v$E, scale=FALSE, organism="Human", top=100)
KO_ctrls <- KO_v$targets$group == "KO_ctrl"
KO_ctrl_mean <- apply(KO_prog[KO_ctrls, ], 2, mean)
KO_ctrl_sd <- apply(KO_prog[KO_ctrls, ], 2, sd)

KO_prog <- t(apply(KO_prog, 1, function(x) x - KO_ctrl_mean))
KO_prog <- apply(KO_prog, 1, function(x) x / KO_ctrl_sd)

# Test using a linear model
## KO severe
KOsev <- KO_v$targets$group %in% c("KO_ctrl", "KO_1")
KOsev_ctrls <- KO_v$targets$group[KOsev] == "KO_ctrl"
KOsev_prog_res <- test_lm(KO_prog[, KOsev], KOsev_ctrls)
print(KOsev_prog_res)
## KO shallow
KOsha <- KO_v$targets$group %in% c("KO_ctrl", "KO_2", "KO_3")
KOsha_ctrls <- KO_v$targets$group[KOsha] == "KO_ctrl"
KOsha_prog_res <- test_lm(KO_prog[, KOsha], KOsha_ctrls)
print(KOsha_prog_res)
```

```{r OE_ssprog}
OE_prog <- progeny(OE_v$E, scale=FALSE, organism="Human", top=100)
OE_ctrls <- OE_v$targets$group == "OE_ctrl"
OE_ctrl_mean <- apply(OE_prog[OE_ctrls, ], 2, mean)
OE_ctrl_sd <- apply(OE_prog[OE_ctrls, ], 2, sd)

OE_prog <- t(apply(OE_prog, 1, function(x) x - OE_ctrl_mean))
OE_prog <- apply(OE_prog, 1, function(x) x / OE_ctrl_sd)

# Test using a linear model
OE_prog_res <- test_lm(OE_prog, OE_ctrls)
print(OE_prog_res)
```

Finally we investigate the sample-wise variability of pathway activities for the records
```{r hp_ssprogeny, fig.width=7, fig.height=4, dpi=300}
mat <- cbind(KO_prog, OE_prog)
gr <- rbind(KO_v$targets[,"group",drop=FALSE],
	    OE_v$targets[,"group", drop=FALSE])
hc <- HeatmapAnnotation(df=gr)
Heatmap(mat, top_annotation = hc, name="Activity\n(scaled to controls)", cluster_columns=FALSE)
```

### PROGENy gene-wide permutation
```{r}
# Get progeny model matrix
progeny.mat <- getModel(organism = "Human", top = 100)
progeny.mat$ID <- rownames(progeny.mat)
progeny.mat <- progeny.mat[, c("ID", setdiff(colnames(progeny.mat), "ID"))]

# Create a data.frame with the universe of genes and corresponding moderated-t
# statistics. Warn: those genes filtered out because of low expression in any
# of the two experiments are considered mot-t = 0, since it is expected that
# both conditions are 0 expresion.
IDs <- unique(c(rownames(KO_eBay$t), rownames(OE_eBay$t)))
df <- data.frame(ID=IDs,
		 KO_severe=rep(0, length(IDs)),
		 KO_shallow=rep(0, length(IDs)),
		 OE=rep(0, length(IDs)),
		 row.names=IDs)
# Fill out with mod-ts
df[rownames(KO_eBay$t), "KO_severe"] <- KO_eBay$t[, "NKD2_KO_severe"]
df[rownames(KO_eBay$t), "KO_shallow"] <- KO_eBay$t[, "NKD2_KO_shallow"]
df[rownames(OE_eBay$t), "OE"] <- OE_eBay$t[, "NKD2_OE"]

# Remove row.names
rownames(df) <- NULL
# Remove duplicated gene symbols
df <- df[which(!duplicated(df$ID)),]

set.seed(1234)
progeny.res <- progenyPerm(df, progeny.mat)
pvals <- apply(progeny.res, 2, function(z) pnorm(z))
pvals <- apply(pvals, c(1,2), function(pval) ifelse(pval > 0.5, (1-pval)*2, pval*2))
fdr <- matrix(p.adjust(as.vector(pvals), method="fdr"), ncol=ncol(pvals), nrow=nrow(pvals),
	      byrow = FALSE, dimnames=dimnames(pvals))

prog <- setNames(vector("list", nrow(progeny.res)), rownames(progeny.res))
for(tag in rownames(progeny.res)) {
    # Make a simple table with the outcome
    progeny.cont <- cbind(Activity=progeny.res[tag,],Pvalue=pvals[tag,], FDR=fdr[tag,])
    # Show in stdout
    cat(paste("Pathway activity from",tag,"\n"), file=stdout())
    print(progeny.cont)
    prog[[tag]] <- progeny.cont
    # Save if as a supplementary table in a file
  write.table(progeny.cont,
              file = paste0(DATADIR,"/",tag,"_progeny.csv"),
              sep=",", col.names = NA, row.names = TRUE)
}
```

```{r make_hp}
### Make a heatmap of the results for the 3 contrasts
# ACT is a matrix of Activity
# PADJ is a matrix of strings for each cell
makeHP <- function(ACT, TXT, cluster_col=FALSE) {

col_fun <- circlize::colorRamp2(c(-3,0,3), c("blue","white","red"))
legend_labs <- c(floor(min(as.vector(ACT))),
		 -2,0,2,
		 ceiling(max(as.vector(ACT))))
hp <- Heatmap(ACT, column_names_side="top", row_names_side="left",
	      col=col_fun, cluster_columns=cluster_col,
	      name="Activity (z-score)",

	      heatmap_legend_param=list(direction="horizontal",
					labels=as.character(legend_labs),
					at=legend_labs,
					labels_gp=gpar(fontsize=12),
					legend_width=unit(6,"cm")),
	show_column_dend=FALSE, show_row_dend=FALSE,	 
 	column_names_rot=0,
	row_names_gp=gpar(fontsize=12),
    	row_names_max_width = max_text_width(
        			rownames(ACT), 
        			gp = gpar(fontsize = 12)),
	column_names_centered=TRUE,
	cell_fun = function(j, i, x, y, width, height, fill) {
		grid.text(paste0(sprintf("%.2f", ACT[i, j]),
				 " (",TXT[i, j],")"), 
			  x, y, 
			  gp = gpar(fontsize = 12,
				    fontface = ifelse(TXT[i,j]!="ns",
						      "bold","plain"),
				    col = ifelse(ACT[i,j] < -1,
						 "white","black"))
			  )
	}
)
return(hp)
}
```

```{r pval2txt}
pval2txt <- function(padj) {
padj_txt <- apply(padj, c(1,2), function(pval) {
	if(pval< 0.001) {
		txt <- "***"
	} else if (pval < 0.01) {
		txt <- "**"
	} else if (pval < 0.05) {
		txt <- "*"
	} else {
		txt <- "ns"
	}
	return(txt)
})

return(padj_txt)
}
```

```{r Fig5k, fig.width=5, fig.height=5, dpi=300}
ACT <- t(progeny.res)
ACT <- ACT[, rev(colnames(ACT))]
PVAL <- t(fdr)
PVAL <- PVAL[, rev(colnames(PVAL))]
stopifnot(all(colnames(ACT)==colnames(PVAL)))

TXT <- pval2txt(PVAL)

hp <- makeHP(ACT, TXT)
draw(hp, heatmap_legend_side="bottom")
```

## Conclusions
> The results obtained by both approaches will be quite similar: JAK-STAT pathway is highly dysregulated towards
NKD2 functional impact in a positive direction. TGFb follows this trend with a strong association too, but lower 
magnitude of dysregulation.

## Save data
```{r}
saveRDS(prog, file=paste0(DATADIR,"/prog.rds"))
```

## Session info

```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```

